 

### 2023-03-02 paper3.Rmd
Revisions to make dplyr::select command run without errors, revisions to Paper 3

list_mental_health_symptoms <- read.xlsx(paste(datapath, 
                                               "list_of_diseases.xlsx", sep = "/")) %>%
  filter(mh_indicator == 1) %>%
  select(search)

@@

{\bf Background:} A key obstacle to studying the effects of public health responses on population mental health in the US is a lack of daily surveillance data on mental health indicators. A potential marker of population mental health is the use of Internet search data, which offers the potential for real-time, large-scale analysis and identification of evolving mental health issues.

{\bf Methods:} We utilize controlled interrupted time series (CITS) approach to evaluate whether the development and announcement of safe and effective COVID-19 vaccines reduced Google searches related to symptoms of depression and anxiety. By comparing how search trends changed in the post period for mental health symptoms above and beyond changes observed in the same period for physical health symptoms, we estimate measures of association between the vaccine announcement and changes in mental health symptom search density. We adjust for state, month, and day of week effects, and we apply a sensitivity analysis to assess the robustness of our estimates to alternative specifications. We further examine whether estimates varied among states and counties that predominantly voted for the Democratic Party candidate in the 2020 presidential election compared to the Republican candidate.

{\bf Findings and interpretation:} Under our main specification, we find evidence of reduced search density associated with the announcement of vaccine safety data for Google search trends for depression and anxiety. `r paper3_abstract`

\emph{Keywords: mental health, controlled interrupted time series, Google searches, United States, COVID-19 vaccines.}

## Introduction
It has been well-documented that throughout the pandemic, symptoms of depression and anxiety have markedly increased among US adults. While it is likely that the pandemic has a direct effect on worsened mental health status due to fear of contracting and spreading the virus, public health responses to the pandemic—including lockdowns, school closures, and social distancing measures—likely have an indirect effect as well. Existing research suggests that the effects of these measures on mental health vary, and may be protective, null, or even harmful. Studies of lockdowns in multiple countries, for example, point to worsened feelings of isolation and hopelessness exacerbated by isolation measures. By contrast, the distribution of safe and effective vaccines against COVID-19 has been shown to reduce symptoms of depression and anxiety by as much as XXX.

* Describe symptoms database and reference published literature using it to predict COVID-19 cases and deaths [@alruilyPredictionCovid19Transmission2022; @abbasAssociationsGoogleSearch2021; @saegnerForecastingSurveillanceCOVID192022]

A recently published evaluation of Google’s COVID-19 Search Trends Symptoms Dataset found strong correlation between search trends for depression and anxiety with indicators of mental health service utilization and need reported in the US Census Household Pulse Survey (HPS) with rates of emergency department visits for mental health condition as reported in the US Centers for Disease Control and Prevention’s (CDC) National Syndromic Surveillance Program (NSSP).[@vaidyanathanEvaluationInternetSearches2022]

controlled interrupted time series (CITS) 
interrupted time series (ITS)

* Motivation to explore vaccine announcement: evidence that vaccination itself reduced depression and anxiety -- do we see earlier reductions from an anticipatory effect?
* Motivation to analyze by polity: Vaccination became politically charged in the US, possible bias from impact of 2020 election on mental health searches
* Overview of approach: CITS using searches for headache as a control, adjusting for fixed effects, subgroup analysis by polity, analysis for 6 different mental health symptoms

### Data

#### Shock date

We take as our first credible signal of the imminent availability of multiple safe and effective COVID-19 vaccines to be November 16, 2020. On November 9, Pzifer announced than early vaccine trial data showed 90% efficacy;[@thomasPfizerEarlyData2020] on the 16th, Moderna similarly announced preliminary results showing 94.5% efficacy for their vaccine.[@gradyEarlyDataShow2020] While U.S. Food and Drug Administration (FDA) Emergency Use Authorization (EUA) was not granted until December, the concurrent announcements of vaccine efficacy were widely reported and interpreted by many public health practitioners, policymakers, and the general public as a substantial signal of a new phase of the pandemic, with Moderna’s chief executive officer calling the news “a game changer” in the fight against COVID-19.

We accessed aggregated national and subnational data from Google’s COVID-19 Search Trends
Symptoms Dataset [@Wahltinez2020] (hereafter referred to as the symptoms dataset) between January 1, 2020 and April 1, 2021. The symptoms dataset provides time series data of search term density, standardized to pairs of health symptoms and regions (e.g., searches for depression in Australia or for fever in Cook County, Illinois) and normalized within each pair based on the symptom’s relative popularity. The time series of search density for each symptom-region pair also contains artificial noise, as Google’s differential privacy technique to data aggregation and anonymization adds random noise to protect user privacy (the errors are symmetrically distributed).

Search density data for over 400 symptoms are reported across national and subnational regions of six countries: Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States. In the United States, subnational regions include states and counties.  Because the aggregated search density trends were constructed based on relative search population within each region, comparison of search density for a specific symptom between different regions is not meaningful; however, comparing search density within regions (or a group of regions) is feasible both over time and across different symptoms.

n_states <- data_its_state %>% select(state_name) %>% unique() %>% nrow()
n_counties <- data_its_county %>% select(county_fips) %>% unique() %>% nrow()

n_states <- format(n_states, big.mark = ",")

For our analysis, we utilized three levels of region from the symptoms dataset: 1) national level data from the six countries in the dataset; 2) U.S. state level data across `r n_states` states; and 3) U.S. county level data reported by `r n_counties` counties. (Though there are over 3,000 counties in the U.S., due to data privacy concerns, smaller county results were not included in the symptoms dataset.) Following the convention of the symptoms dataset, we will refer to the geographic units (e.g., states, counties, and nations) within the data as "regions."

We further obtained county-level presidential election returns for 2020 from the MIT Election Data and Science Lab,[@mitelectiondataandsciencelabCountyPresidentialElection2022] which we then used to calculate aggregate county and state votes (including absentee, early, and election day ballots) won by each candidate. The aggregate votes were then used to determine vote share and margins (e.g., the percentage points difference in vote share between the candidate with the most votes to the runner up). These margins were then used to sort counties and states into quintiles, such that the 20% most Republican-voting localities appearing in the symptoms dataset were sorted into the first quintile and the 20% most Democratic-voting localities sorted into the fifth).

Population estimates for 2020 by Federal Information Processing Standards (FIPS) code were obtained from the United States Department of Agriculture (USDA) Economic Research Service.

All analyses were completed using R (version 4.2.1).[@rcoreteamLanguageEnvironmentStatistical2022]

We use an ITS approach to model Google search density for a given symptom $d$ using the following baseline equation: $$Y_{d,g, i} = \lambda_0 + \lambda_1 time_t + \alpha_d Post_{d,i} + \beta_d time_t*post_{d,i} + \theta_r + T^{'}_t\chi + \epsilon_{it} \ \ \ ,\ \ \ \ \ \ \ \ \ \ \ \ (3.1)$$ where $time_t$ is a continuous variable that captures the days leading to the vaccine announcement, $Post_i$ is an indicator variable which equals 1 in the period after the shock data and 0 otherwise, $\theta_r$ is a fixed effect for region $r$, and $T^{'}_t\chi$ is a matrix of time-vary fixed effects (i.e., month and day of week).

In our model, $\alpha_d$ can be interpreted as a level or intercept associated with the shock date—that is, the immediate difference in search density for a disease $d$ following November 19, 2020. The term $\beta_d$—i.e., the slope change—can be interpreted as the sustained change in search density in the days following November 16th.

```
We apply this model for all available `r diseases_n` diseases with county-level data. As search density may rapidly change in response to a shock, we trimmed the bandwidth of dates to fit our model to forty days before and after November 16, 2020.

To control for time-varying confounders which may have affected symptom search trends during the study period, we applied a CITS approach to contrast how searches for mental health symptoms varied above and beyond changes in search density for a comparable physical complaint (i.e., headaches).

After reviewing the list of symptoms for which search density data were available, we identified the following `r mh_diseases_n` mental health related symptoms: `r paste0(gsub("asperger", "Asperger", gsub("_", " ", printList(toPrint = sort(list_mental_health_symptoms), finalSepWord = "and", midSep = ","))),".")`

For each of these mental health symptoms, $d$, we ran the following model to predict search density:

$$Y_{d,it} = \lambda_0 + \lambda_1 time_t + \lambda_2 Post_t + \lambda_3 time_t*Post_t + \alpha_d treated_{di} + 
\ \ \ \ \ $$
$$\beta_d time_t*treated_{di}  + \delta_{d} Post_i*treated_{di} + \gamma_d time_t*Post_i*treated_{di} + \theta_g + T^{'}_t\chi +  \epsilon_{it}\ \ \ ,\ \ \ \ \ \ \ \ \ \ (3.2)$$
\end{spacing}

where $treated_{di}$: is an indicator variable which equals 1 if the observation is for a mental health condition $d$, 0 if for headaches. The CITS effect estimates of interest are given by $\delta$, which reflects the additional level shift among mental health related symptoms searches compared to searches for headaches after the vaccine announcement, and  $\gamma$, the additional change in search density over time (i.e., slope change) comparing the same.

Given the high number of observations in our sample, we evaluated significance against an alpha of 0.01, and we constructed 99% confidence intervals using heteroskedasticity-robust standard errors.

As with other quasi-experimental methods, it is recommended that the modeling choices and specifications of a CITS analysis be evaluated using a sensitivity analysis. To do so, we defined a list of seven methodological choices concerning data processing and modeling assumptions and interacted these choices to generate all their plausible  combinations. These combinations were then conducted to allow for a multiverse analysis to identify how sensitive effect estimates and confidence intervals were to different methodological choices.

These criteria used to generate the plausible model combinations are outlined in Table 3.1.

First, we considered the role of data cleaning in the outcome data, which involved outlier smoothing using Friedman’s nonparametric super smoother regression estimator.[@friedmanVariableSpanScatterplot1984] Second, we compared the level of region of aggregated data, using U.S. counties, U.S. states, and, finally, national trends from the six counties in the sample. Third, we evaluated sensitivity to our specification of the bandwidth cut-off around the shock date, extending this from ±40 days to 40 and 60 days. Fourth, we toggled whether to account for day of week fixed effects. Fifth, we toggled month fixed effects. Sixth, we similarly toggled regional fixed effects (i.e., fixed effects for states in the U.S. state level data, fixed effects for counties in the U.S. county level data, etc.).

Interacting these choices resulted in 144 plausible models for each symptom. Each of the 144 models were run for all `r mh_diseases_n` mental health related symptoms, resulting in `r num_models_multiverse` models being conducted overall.

The first subgroup analysis applied the CITS approach against all `r mh_diseases_n` mental health related symptoms against search data from each of the six countries in the sample, using our nationally aggregated data. This analysis aimed to investigate the extent to which effects observed in US states and counties were also seen in other countries.

The second subgroup analysis applied the CITS approach to the 20% most Republican-voting localities (hereafter referred to as red states/counties) and to the 20% most Democratic-voting localities (hereafter referred to as blue states/counties) based on 2020 presidential election vote share. A growing body of evidence has documented that political ideology can play a role in how receptive people are to vaccination; this has been highlighted throughout the COVID-19 pandemic, with people identifying as conservative being more likely to express skepticism of public health authorities and the scientific community and being more likely to express concerns about the safety, effectiveness, or necessity of vaccines. Our subgroup analysis comparing CITS effect estimates in blue vs. red states aimed to investigate whether the initial response to imminent safe and effective COVID-19 vaccines showed evidence of heterogeneity by predominant political ideology, under the hypothesis that if conservative voters were more likely to be averse to vaccination, any impact on mental health related symptom search density from the announcement of upcoming vaccines would be lesser in red states and counties compared to blue ones.

The Institutional Review Board (IRB) of the Harvard T.H. Chan School of Public Health has determined that the study was not human subjects research, and that additional review was not required (protocol number: IRB23-0095, determined on January 25, 2023).

@@

function_dynamic_text_results <-
  function(section = "its",         #c(its, main, international, polity, multiverse)
           disease_option,
           coefficient_option,
           lag_option,
           region_option = "all",
           geography_option = "county",
           value_include = TRUE,
           multiverse_sign = "positive",
           multiverse_p = FALSE) #(TRUE, FALSE, only),)

@@

    val <- round((output_df$estimate), 3)
    p_val <- round(output_df$p_value, 3)
    low <- round((output_df$lower_99), 3)
    high <- round((output_df$upper_99), 3)

@@

      return(paste0(
        val,
        " percentage points",
        ifelse(coefficient_option == "slope",
               " per day "," "),
        "(p ",
        ifelse(p_val < 0.001,
               "< 0.0001, ",
               paste0("= ",
                      p_val,
                      ", ")),
        "99\\% CI: ",
        low,
        " – ",
        high
        ,
        ")"))

@@

@@

      plot <- data_its_temp %>% 
        filter(days_post_shock %in% (-40:40) ) %>%
        mutate(trt = factor(trt, 
                            levels = c(m2,
                                       "Headache"))) %>%
        ggplot(aes(x = days_post_shock, y = fitted, group = trt, color = trt)) +
        geom_boxplot(aes(x = days_post_shock, y = daily_mean_clean, 
                         group = interaction(trt, days_post_shock), color = trt, 
                         fill = trt, outlier.shape = NA),
                     outlier.shape = NA,
                     alpha = 0.01) +
        geom_line() +
        theme_classic() +
        labs(x = "Days after November 16, 2020",
             y = "Density of Google Search trends",
             color = "",
             title = ifelse(g == "state", "US states", 
                            ifelse(g == "county", "US counties",
                                   "International data"))) +
        theme(legend.position = "bottom") +
        geom_vline(xintercept = 0, linetype='dotted') +
        scale_color_manual(values = c(col_mh, col_headache),
                           name = "") +
        scale_fill_manual(values = c(col_mh, col_headache),
                          name = "")

@@

@@

The ITS estimates for the immediate (level shift) and sustained (slope change) effects associated with the announcement of vaccine safety and efficacy data on search density for all `r diseases_n` are presented in Figure \ref{figure_all_diseases}.

@@

```{r figure_all_diseases, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 6, fig.height = 4.5, fig.align='center', out.width = '95%',  fig.scap="Interrupted time series (ITS) effect estimates and confidence intervals by symptom.", out.extra='',  fig.cap="\\label{figure_all_diseases}Interrupted time series (ITS) effect estimates and confidence intervals by symptom. Effect estimates are presented by points;  for each symptom, the estimated slope change per day is shown on the x-axis, while the level shift over the post-announcement period is shown on the y-axis. The horizontal and vertical bars next to each point represent the 99\\% confidence intervals for the estimate slope change and level shift, respectively. Mental health-related symptoms are shown in blue, while physical symptoms are shown in grey. ADHD: Attention deficit hyperactivity disorder. MDD: Major depressive disorder."}

model_outputs_its_graph <- model_outputs_its %>%
  mutate(coefficient = ifelse(coefficient == "post:time", "slope",
                              ifelse(coefficient == "post", "level", "ERROR"))) %>%
  select(coefficient, estimate, lower_95, upper_95, disease, p_value) %>%
  pivot_wider(names_from = coefficient,
              values_from = c(estimate, lower_95, upper_95, p_value)) %>%
  mutate(mh_indicator = ifelse(disease %in% list_mental_health_symptoms, 
                               "Mental symptom",
                               ifelse(disease == "headache", 
                                      "Headache (control)",
                                      "Physical symptom")))  %>%
  mutate(mh_indicator = factor(mh_indicator, 
                               levels = c("Mental symptom", 
                                          "Headache (control)",
                                          "Physical symptom"))) %>%
  function_clean_diseases()

@@

# Make tables
# Run international comparison models
# Generate in text formulas

>>>>

list_mental_health_symptoms <- read.xlsx(paste(datapath, 
                                               "list_of_diseases.xlsx", sep = "/")) %>%
  filter(mh_indicator == 1) %>%
  dplyr::select(search)

@@

@@

function_dynamic_text_results <-
  function(section = "its",         #c(its, main, international, polity, multiverse)
           disease_option,
           coefficient_option,
           abstract_option = FALSE,
           abstract_option_CI = FALSE,
           region_option = "all",
           geography_option = "county",
           value_include = TRUE,
           multiverse_sign = "positive",
           multiverse_p = FALSE) #(TRUE, FALSE, only),)

@@

    val <- format(round((output_df$estimate), 3), nsmall = 3)
    p_val <- format(round(output_df$p_value, 3), nsmall = 3)
    low <- format(round((output_df$lower_99), 3), nsmall = 3)
    high <- format(round((output_df$upper_99), 3), nsmall = 3)

@@

      if(abstract_option == TRUE){

        return(paste0(
        val,
        " percentage points",
        ifelse(coefficient_option == "slope",
               " per day "," "),
        "(99\\% ",
        ifelse(abstract_option_CI == TRUE, "confidence interval (CI): ",
               "CI: "),
        low,
        " to ",
        high
        , 
        "; p ",
        ifelse(p_val < 0.001,
               "< 0.0001)",
               paste0("= ",
                      p_val,
                      ")"))))

@@

      else {
      return(paste0(
        val,
        " percentage points",
        ifelse(coefficient_option == "slope",
               " per day "," "),
        "(p ",
        ifelse(p_val < 0.001,
               "< 0.0001, ",
               paste0("= ",
                      p_val,
                      ", ")),
        "99\\% CI: ",
        low,
        " – ",
        high
        ,
        ")"))

      }
    }

@@

      plot <- data_its_temp %>% 
        filter(days_post_shock %in% (-40:40) ) %>%
        mutate(trt = factor(trt, 
                            levels = c(m2,
                                       "Headache"))) %>%
        ggplot(aes(x = days_post_shock, y = fitted, group = trt, color = trt)) +
        geom_boxplot(aes(x = days_post_shock, y = daily_mean_clean, 
                         group = interaction(trt, days_post_shock), color = trt, 
                         fill = trt, outlier.shape = NA),
                     outlier.shape = NA,
                     alpha = 0.01) +
        geom_line() +
        theme_classic() +
        labs(x = "Days after November 16, 2020",
             y = "Density of Google Search trends",
             color = "",
             title = ifelse(g == "state", "U.S. states", 
                            ifelse(g == "county", "U.S. counties",
                                   "International data"))) +
        theme(legend.position = "bottom") +
        geom_vline(xintercept = 0, linetype='dotted') +
        scale_color_manual(values = c(col_mh, col_headache),
                           name = "") +
        scale_fill_manual(values = c(col_mh, col_headache),
                          name = "")

@@

paper3_abstract <- paste0("For searches related to depression, we observe a sustained decline in search density of ",
                          function_dynamic_text_results("main","depression", "slope" , TRUE, TRUE), " and an immediate downward level shift of ",
                          function_dynamic_text_results("main","depression", "intercept", TRUE),
                          " relative to comparable symptom searches. For anxiety, we estimated a similar slope change of",
                          function_dynamic_text_results("main","anxiety", "slope", TRUE), 
                          " and a level shift of ",
                          function_dynamic_text_results("main","anxiety", "intercept", TRUE),
 ". We did not observe significant variations across predominately Democrat or Republican voting states and counties, and our findings were robust to alternative specifications of bandwidth, fixed effects, and data cleaning decisions.")

@@

{\bf Background:} A key obstacle to studying the impact of policies on population mental health in the U.S. is a lack of appropriate high frequency surveillance data. A potential marker of population mental health is the use of Internet search data, which enables real-time, large-scale analysis of evolving mental health issues.

{\bf Methods:} We utilized controlled interrupted time series approaches to evaluate whether the development of COVID-19 vaccines reduced Google searches related to symptoms of depression and anxiety in the U.S.. We examined how search trends changed following the announcement of vaccine safety and efficacy data for mental health symptoms compared with physical health symptoms, and estimated associations between vaccine announcement and changes in mental health symptom search density. We adjusted for state, month, and weekday effects, and applied numerous sensitivity analyses to assess the robustness of our findings. To explore heterogeneity by predominate political affiliation, we also tested whether our estimates varied among states and counties that predominantly voted for the Democratic Party candidate in the 2020 presidential election compared to the Republican candidate.

{\bf Findings and interpretation:} Under our main model specification, we observed reduced search density for depression and anxiety associated with the announcement of vaccine safety data. `r paper3_abstract` If symptom search density echoes population health, our results would indicate that announcements regarding the development of safe and effective COVID-19 vaccines led to a robust and sustained improvement in mental health. Our findings underscore the importance of timely and transparent communication and highlight the potential application of high frequency internet search data for population mental health surveillance.

\emph{Keywords: mental health; controlled interrupted time series; COVID-19 vaccines; United States; Google searches.}

## Introduction
The COVID-19 pandemic has had a profound effect on the mental health and wellbeing of individuals worldwide. Studies from Australia,[@batterhamTrajectoriesDepressionAnxiety2021] China,[@ahmedEpidemicCOVID19China2020; @gaoMentalHealthProblems2020; @hwangLonelinessSocialIsolation; @leiComparisonPrevalenceAssociated2020; @liuPrevalencePredictorsPTSS2020; @qiuNationwideSurveyPsychological2020; @wangImmediatePsychologicalResponses2020; @wangStudyPublicPsychological2021; @wangPsychologicalDistressCoping2020; @zhangImpactCOVID19Pandemic2020] Denmark,[@sonderskovDepressiveStateDenmark] Europe,[@hajekPrevalenceDeterminantsProbable2022] Iran,[@moghanibashi-mansouriehAssessingAnxietyLevel2020] Ireland,[@hylandAnxietyDepressionRepublic2020]  Italy,[@mazzaNationwideSurveyPsychological2020; @mocciaAffectiveTemperamentAttachment2020] Nepal,[@samadarshiOnlineSurveyFactors2020] New Zealand,[@gasteigerDepressionAnxietyStress2021; @lawRiskFactorsChanges2022] Spain,[@gonzalez-sanguinoMentalHealthConsequences2020; @ozamiz-etxebarriaStressAnxietyDepression2020] Turkey[@ozdinLevelsPredictorsAnxiety2020],  the United States,[@jiaNationalStateTrends2021, p. 2020–2021; @kujawaExposureCOVID19Pandemic2020] and the United Kingdom[@dettmannInvestigatingPrevalenceAnxiety2022] have reported high levels of anxiety (6-51%), depression (15-48%), and post-traumatic stress disorder (7-54%) associated with the onset of the pandemic. In the United States, approximately four in ten adults reported experiencing symptoms of anxiety or depression in December 2020—a four-fold increase from January-June 2019.[@panchalImplicationsCOVID19Mental2021, p. 19] Globally, it has been estimated that the prevalence of depression and anxiety during the COVID-19 pandemic may be as high as 24% and 21%, respectively.[@banko-ferranImpactCOVID19Mental2020; @castaldelli-maiaInvestigatingEffectNational2021] This striking burden has been described as a mental health crisis[@melimopoulosHowCOVID19Pandemic2021] and as a pandemic in its own right.[@kwaiWhatPointYoung2021]

While COVID-19 continues to bring unprecedented challenges to societies around the world, the development and distribution of safe and effective vaccines has fundamentally changed the trajectory of the pandemic. Beyond the physical benefits vaccines provide in reducing the risk of infection and associated morbidity and mortality, a growing body of evidence suggests that vaccination has also
contributed to improved mental health.[@agrawalImpactCOVID19Vaccine2021; @chourpiliadisShorttermImprovementMental2023; @nguyenPsychologicalBenefitsCOVID192021; @rubinsteinEffectCOVID19Vaccinations2023] While there is limited international data on the secondary mental health impact of COVID-19 vaccination, research drawing from the U.S. Census Bureau’s Household Pulse Survey has found that benefits of vaccination may include up to a 30% reduction in symptoms of anxiety and depression among vaccinated adults,[@agrawalImpactCOVID19Vaccine2021] while research using the University of Southern California Understanding America Survey suggests a reduction in loneliness associated with greater social interactions following vaccination.[@leeEffectVaccinationSocial2022]

While this literature suggests that receiving a COVID-19 vaccine can improve mental health, the potential mental health benefits of announcements regarding the development of vaccines have received comparatively less attention.[@cotfasLongestMonthAnalyzing2021; @kesselEffectCOVID19Pandemic2023] In the context of a deadly pandemic, the announcement of a safe and effective vaccine could be a powerful signal, providing a psychological boost that the worst might soon be mitigated. Studying how population mental health may have been impacted by the release of vaccine efficacy and safety trial data may provide insights into how emotional well-being changed throughout the pandemic in response to these signals, which, in turn, could inform public health communication and contribute to our understanding of how the public responds to information about vaccines—a necessary understanding for increasing vaccine uptake and trust in public health responses.

A key obstacle to studying the impact of public health policies on population mental health, however, is a lack of appropriate high frequency surveillance data. A potential marker of population mental health is the use of internet search data, which enables real-time, large-scale analysis of evolving mental health issues. A recently published evaluation of Google’s COVID-19 Search Trends Symptoms Dataset[@Wahltinez2020] found a strong correlation between search trends for depression and anxiety with indicators of mental health services utilization (reported in the U.S. Census Household Pulse Survey) and rates of emergency department visits for mental health conditions (from the U.S. Centers for Disease Control and Prevention’s (CDC) National Syndromic Surveillance Program).[@vaidyanathanEvaluationInternetSearches2022] If taken as a reasonable proxy for mental health disorders, search data can provide valuable information on how population mental health reacts to real-time events, enabling policymakers to better respond to mental health needs in a timely manner.

Our study analyzes whether news announcements (in November 2020) regarding the imminent availability of safe and effective COVID-19 vaccines impacted population mental health—even before these vaccines were made available to the public. Using Google search density data, we applied interrupted time series approaches to investigate how these announcements might have affected search trends for mental health symptoms in the U.S. and five other countries where data were available: Australia, the United Kingdom, Ireland, New Zealand, Singapore. Owing to emerging evidence that political ideology may be a predictor of confidence in public health and vaccine hesitancy,[@debusPoliticalIdeologyVaccination2021; @baumgaertnerInfluencePoliticalIdeology2018; @stolerSociopoliticalPsychologicalCorrelates2022] we further explored whether changes in search trends for mental health symptoms varied across U.S. states and counties that predominantly voted Democrat vs. Republican in the 2020 presidential election. Our findings suggest declines in density for searches related to anxiety and depression in all of the countries examined; in the U.S., these declines were found to be consistent irrespective of political ideology.

### Data sources

#### Vaccine announcements

We take as our first credible signal of the imminent availability of safe and effective COVID-19 vaccines to be November 16, 2020. On November 9^th^, Pzifer announced that early vaccine trial data showed 90% efficacy[@thomasPfizerEarlyData2020] and on the 16^th^, Moderna announced preliminary results showing 95% efficacy.[@gradyEarlyDataShow2020] While U.S. Food and Drug Administration (FDA) Emergency Use Authorization (EUA) was not granted until December, the concurrent announcements of vaccine efficacy were widely reported and interpreted by many public health practitioners, policymakers, and the general public as a substantial signal of a new phase of the pandemic, with Moderna’s chief executive officer calling the news “a game changer” in the fight against COVID-19.

We accessed aggregated national and subnational data from Google’s COVID-19 Search Trends Symptoms Dataset [@Wahltinez2020] (hereafter referred to as the symptoms dataset) between January 1, 2020 and April 1, 2021. The symptoms dataset provides time series data of search term density, standardized to pairs of health symptoms and regions (e.g., searches for depression in Australia or fever in Cook County, Illinois) and normalized within each pair based on the symptom’s relative popularity. The time series of search density for each symptom-region pair also contains artificial noise, as Google’s differential privacy technique to data aggregation and anonymization adds random noise to protect user privacy (the errors are symmetrically distributed).[@Wahltinez2020]

Search density data for over 400 symptoms were reported by Google across national and subnational regions of six predominately English-speaking countries: Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the U.S. In the U.S., subnational regions include states and counties.  Because the aggregated search density data were constructed based on relative search population within each region, we could not compare search density for specific symptoms across different regions; however, comparing search density within regions (or a group of regions) was feasible both over time and across different symptoms.

n_states <- data_its_state %>% dplyr::select(state_name) %>% unique() %>% nrow()
n_counties <- data_its_county %>% dplyr::select(county_fips) %>% unique() %>% nrow()

n_states <- format(n_states - 1, big.mark = ",")

We utilized three levels of region: 1) national level for the six countries; 2) U.S. state level data across `r n_states` states and the District of Colombia; and 3) U.S. county level across `r n_counties` counties (though there are over 3,000 counties in the U.S., due to data privacy concerns, smaller county results were not included in the symptoms dataset). Following the convention of the symptoms dataset, we will refer to the geographic units (e.g., states, counties, and nations) within the data as "regions."

We obtained county-level presidential election results for 2020 from the MIT Election Data and Science Lab,[@mitelectiondataandsciencelabCountyPresidentialElection2022] which we used to calculate aggregate county and state votes (including absentee, early, and election day ballots) for each candidate. The aggregate votes were then used to determine vote share and margins (e.g., the percentage points difference in vote share between the winning candidate and the runner up). Applying a similar approach as other research into COVID-19 and political party inclination,[@grossmannAllStatesClose2021] we used these margins to sort counties and states into quintiles, such that the 20% most Republican-voting localities appearing in the symptoms dataset were sorted into the first quintile and the 20% most Democratic-voting localities sorted into the fifth quintile.

Population estimates for 2020 by Federal Information Processing Standards (FIPS) code were obtained from the U.S. Department of Agriculture (USDA) Economic Research Service.

All analyses were conducted using R software (version 4.2.1).[@rcoreteamLanguageEnvironmentStatistical2022]

We used an interrupted time series (ITS) approach to model the Google search density for a given symptom $d$ using the following baseline equation: $$Y_{d,g, i} = \lambda_0 + \lambda_1 time_t + \alpha_d Post_{d,i} + \beta_d time_t*post_{d,i} + \theta_r + T^{'}_t\chi + \epsilon_{it} \ \ \ ,\ \ \ \ \ \ \ \ \ \ \ \ (3.1)$$ where $time_t$ is a continuous variable that captures the days before the vaccine announcement ($time_0$), $Post_i$ is an indicator variable equal to 1 in the period after the announcement (0 otherwise), $\theta_r$ indicates a series of fixed effects for each region $(r)$, $T^{'}_t\chi$ is a matrix of time fixed effects (i.e., month and day of week), and $\epsilon_{it}$ is an error term. $\alpha_d$ can be interpreted as a level change associated with the announcement date—that is, the immediate difference in search density for a disease $d$ following November 19, 2020; $\beta_d$—the slope change—can be interpreted as the sustained change in search density in the days following November 16th.

```
We apply this model for all available `r diseases_n` diseases with county-level data. As search density may rapidly change in response to new announcements, we trimmed the bandwidth of dates to fit our model to 40 days before and after November 16, 2020.

To control for time-varying confounders which may affect symptom search trends, we applied a controlled interrupted time series (CITS) approach to contrast how searches for mental health symptoms varied above and beyond changes in search density for a comparable physical complaint (i.e., headaches).

After reviewing the list of symptoms for which search density data were available, we identified the following `r mh_diseases_n` mental health symptoms: `r paste0(gsub("asperger", "Asperger", gsub("_", " ", printList(toPrint = sort(list_mental_health_symptoms), finalSepWord = "and", midSep = ","))),".")`

For each of these symptoms, $d$, we ran the following model:

$$Y_{d,it} = \lambda_0 + \lambda_1 time_t + \lambda_2 Post_t + \lambda_3 time_t*Post_t + \alpha_d mhealth_{di} + 
\ \ \ \ \ $$
$$\beta_d time_t*mhealth_{di}  + \delta_{d} Post_i*mhealth_{di} + \gamma_d time_t*Post_i*mhealth_{di} + \theta_g + T^{'}_t\chi +  \epsilon_{it}\ \ \ ,\ \ \ \ \ \ \ \ \ \ (3.2)$$
\end{spacing}

where $mhealth_{di}$ equals 1 if the observation is for a mental health symptom $d$ (0 for headaches). The estimates of interest are given by $\delta$, which reflects the additional level shift among symptom searches (compared to searches for headaches) after vaccine announcement, and  $\gamma$, the additional relative change in search density over time (i.e., slope change).

Given the high number of observations in our sample, we evaluated significance against an alpha of 0.01, and we constructed 99% confidence intervals (CIs) using heteroskedasticity-robust standard errors.

As with other quasi-experimental methods, it is recommended that the modeling specifications be evaluated using a sensitivity analysis. To do so, we defined a list of seven methodological choices concerning data processing and modeling assumptions and interacted these choices to generate all plausible combinations. These combinations were then conducted to allow for a multiverse analysis to identify how sensitive effect estimates and CIs were to different methodological choices. The criteria used to generate the plausible model combinations are outlined in Table 3.1.

First, we considered the role of data cleaning in the outcome data, which involved outlier smoothing using Friedman’s nonparametric super smoother regression estimator.[@friedmanVariableSpanScatterplot1984] Second, we compared the level of region of aggregated data, using U.S. counties, U.S. states, and, finally, national trends from the six counties in the sample. Third, we evaluated sensitivity to our specification of the bandwidth cut-off around the announcement date, extending it from ±40 days to ±50 and ±60 days. Fourth, we toggled whether to account for day of week fixed effects; fifth, we toggled month fixed effects; and sixth, we toggled regional fixed effects (i.e., fixed effects for states in the U.S. state level data, fixed effects for counties in the U.S. county level data, etc.).

Interacting these choices resulted in 144 plausible models for each symptom. Each model was run for all `r mh_diseases_n` mental health related symptoms, resulting in `r num_models_multiverse` models being conducted overall.

The first subgroup analysis applied the CITS approach against all `r mh_diseases_n` mental health related symptoms against search data from each of the six countries in the sample, using our nationally aggregated data. This analysis aimed to investigate the extent to which effects observed in U.S. states and counties were also seen in other countries.

The second subgroup analysis applied the CITS approach to the 20% most Republican-voting localities (hereafter referred to as red states/counties) and to the 20% most Democratic-voting localities (hereafter referred to as blue states/counties) based on 2020 presidential election vote share. Our subgroup analysis comparing CITS effect estimates in blue vs. red states aimed to investigate whether the initial response to imminent safe and effective COVID-19 vaccines showed evidence of heterogeneity by predominant political ideology, under the hypothesis that if conservative voters were more likely to be averse to vaccination, any impact on mental health related symptom search density from the announcement of upcoming vaccines would be lesser in red states and counties compared to blue ones.

The Institutional Review Board (IRB) of the Harvard T.H. Chan School of Public Health determined that the study was not human subjects research, and that additional review was not required (protocol number: IRB23-0095, determined on January 25, 2023).

The estimates for the immediate (level shift) and sustained (slope change) effects associated with the announcement of vaccine safety and efficacy data on search density for all `r diseases_n` are presented in Figure \ref{figure_all_diseases}.

@@

```{r figure_all_diseases, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 6, fig.height = 4.5, fig.align='center', out.width = '95%',  fig.scap="Interrupted time series effect estimates and confidence intervals by symptom.", out.extra='',  fig.cap="\\label{figure_all_diseases}Interrupted time series effect estimates and confidence intervals by symptom. Effect estimates are presented by points;  for each symptom, the estimated slope change per day is shown on the x-axis, while the level shift over the post-announcement period is shown on the y-axis. The horizontal and vertical bars next to each point represent the 99\\% confidence intervals for the estimate slope change and level shift, respectively. Mental health-related symptoms are shown in blue, while physical symptoms are shown in grey. ADHD: Attention deficit hyperactivity disorder. MDD: Major depressive disorder."}

model_outputs_its_graph <- model_outputs_its %>%
  mutate(coefficient = ifelse(coefficient == "post:time", "slope",
                              ifelse(coefficient == "post", "level", "ERROR"))) %>%
  dplyr::select(coefficient, estimate, lower_95, upper_95, disease, p_value) %>%
  pivot_wider(names_from = coefficient,
              values_from = c(estimate, lower_95, upper_95, p_value)) %>%
  mutate(mh_indicator = ifelse(disease %in% list_mental_health_symptoms, 
                               "Mental symptom",
                               ifelse(disease == "headache", 
                                      "Headache (control)",
                                      "Physical symptom")))  %>%
  mutate(mh_indicator = factor(mh_indicator, 
                               levels = c("Mental symptom", 
                                          "Headache (control)",
                                          "Physical symptom"))) %>%
  function_clean_diseases()

@@

# Make tables
